Lens Recovery with GLASS: An Extensible Free-Form Gravitational Lens Modeling Framework

### 2013-05-28 limits/ms.tex
New text describing the features of Glass. Additional plotting macro.

% From aastex.cls
\newcommand\plotone[1]{%
 \centering
 \leavevmode
 \includegraphics[width={\columnwidth}]{#1}%
}%
\newcommand\plottwo[2]{{%
 \centering
 \leavevmode
 \columnwidth=.45\columnwidth
 \includegraphics[width={\columnwidth}]{#1}%
 \hfil
 \includegraphics[width={\columnwidth}]{#2}%
}}%

@@

\Glass\ is an evolution of the concepts employed by the free form modeling tool
\PixeLens\ but completely rewritten in Python and C. The key improvements are:
\begin{enumerate}
  \item A modular framework allows new priors to be added and modified easily.
  \item The basis functions approximating a model can be changed. 
  \item Non-linear constraints can be imposed in an automated post-processing step. 
  \item The central region can be have a higher resolution to capture steep models. 
  \item Stellar density can be used as an additional constraint.  
  \item Point or extended mass objects can be placed in the field.
  \item Additional levels of pixel refinement.
\end{enumerate}
At the heart of \Glass\ lies a new algorithm for sampling the high dimensional
linear space that represents the modeling solution space. This algorithm was
described and tested in \cite{}. \Glass\ is also multi-threaded allowing it to
run efficiently on many-cored machines.  The software is freely available for
download. One interesting aspect is that the input files are themselves Python
programs. This allows a large amount of sophistication in setting up a lens at
runtime and also allows \Glass\ to be used as an external library from another
user written program.

@@

The pixels only cover a finite circular area with physical radius $\Rmap$ and
pixel radius $\Rpix$ centered on the lensing galaxy. To account for any global
shearing outside this region from, e.g., a neighboring galaxy, we also add to
\eqnref{discrete potential} two shearing terms
%
\begin{equation}
\gamma_1(\theta_x^2 - \theta_y^2) + 2\gamma_2\theta_x\theta_y\quad.
\end{equation}
%
We can continue adding terms to account for assumed potentials or potentials
induced from the presence of other galaxy in the field. If the stellar density
has been estimated we can use this as a lower bound where the stellar potential
is a known constant of the form \eqnref{discrete potential}.
%
%\begin{equation}
  %\psi^*(\vec\theta) = \sum_n \kappa^*_n Q_n(\vec\theta)\quad.
%\end{equation}
%

@@

\begin{table}
\begin{tabular}{cllllllll}
Galaxy & $\gamma_\star$ & $M_\star$ & $\gamma_\mathrm{DM}$ & $M_\mathrm{DM}$ & $\Rmap$ & Notes\\
\hline
AA & 1 & 4 & 0.05 & $11^{2.95}$ & 50 kpc & \\
AC & 1 & 4 & 1 & $11^2$ & 50 kpc & \\
BB & 1.5 & $2^{1.5}$ & 0.16 & $11^{2.84}$ & 50 kpc & \\
BC & 1.5 & $2^{1.5}$ & 1 & $11^2$ & 10 kpc & 
\end{tabular}
\caption{Profile parameters for the four mock galaxies. The resulting profiles only roughly follow
\eqnref{Dehnen profile} because the galaxies are triaxial. Masses are in units of $1.8\e{10}\Msun$. The scale lengths for
all lenses are $(a_\star,a_\mathrm{DM})=(2,20)$ kpc and by definition
$M_\star(a_\star) = M_\mathrm{DM}(a_\star) = 1$.}
\label{mock galaxy params}
\end{table}

\begin{figure}
\plotone{cmp_simulations.pdf}
\caption{
\textbf{(Left)} 
Three-dimensional density of the galaxies showing the stellar, dark matter,
and total densities with dotted, dashed, and solid lines, respectively. The
grey lines show the analytic forms from \eqnref{Dehnen profile}. 
\textbf{(Middle)} 
The radially averaged two-dimensional projected density in terms of the
critical density for lensing. At $z_L=0.31$, $\kappa_\mathrm{crit}\sim 1.8\e{9}$\Msun/kpc$^2$.
\textbf{(Right)}
The enclosed projected mass.
}
\label{mock galaxies}
\end{figure}

@@

\subsection{Stellar mass}

>>>>

% From aastex.cls
\newcommand\plotone[1]{%
 \centering
 \leavevmode
 \includegraphics[width={\columnwidth}]{#1}%
}%
\newcommand\plottwo[2]{{%
 \centering
 \leavevmode
 \columnwidth=.45\columnwidth
 \includegraphics[width={\columnwidth}]{#1}%
 \hfil
 \includegraphics[width={\columnwidth}]{#2}%
}}%
\newcommand\plotthree[3]{{%
 \centering
 \leavevmode
 \columnwidth=.30\columnwidth
 \includegraphics[width={\columnwidth}]{#1}%
 \hfil
 \includegraphics[width={\columnwidth}]{#2}%
 \hfil
 \includegraphics[width={\columnwidth}]{#3}%
}}%

@@

\Glass\ is an evolution of the concepts employed by the free form modeling tool
\PixeLens\ but completely rewritten in Python and C. The key improvements are:
%
\begin{enumerate}
  \setcounter{enumi}{0}
  \item A modular framework allows new priors to be added and modified easily.
\end{enumerate}
%
Each prior is a simple function that adds linear constraints that operate on either
a single lens object or the entire ensemble of objects. \Glass\ comes with a number
of useful priors, but a user can write their own in the input file.
%
\begin{enumerate}
  \setcounter{enumi}{1}
  \item The basis functions approximating a model can be changed. 
\end{enumerate}
%
\Glass\ currently describes the lens mass as a collection of pixels, but the code
has been designed to support alternative methods. In particular, it is planned
to develop a module using Bessel functions. This will require a new set of 
priors that operate on these functions.
%
\begin{enumerate}
  \setcounter{enumi}{2}
  \item Non-linear constraints can be imposed in an automated post-processing step. 
\end{enumerate}
%
Once \Glass\ has generated an ensemble of models given the linear contraints, any number
of post processing functions can be applied. Not only can these functions be used to
derive new quanities from the mass models, they can also be used as a filter to 
accept or reject a model based on some non-linear constraint. The plotting functions
within \Glass\ will correctly display models that have been accepted or rejected.
%
\begin{enumerate}
  \setcounter{enumi}{3}
  \item The central region can be have a higher resolution to capture steep models. 
\end{enumerate}
%
By default, the mass distribution of the lens is described by uniform grid. However,
in the central region of a lensing galaxy where the mass profile may rise steeply,
a higher resolution can be specified that only applies to the inner pixels. These
pixels are subdivided into a smaller grid.
%
\begin{enumerate}
  \setcounter{enumi}{4}
  \item Stellar density can be used as an additional constraint.  
\end{enumerate}
%
The mass in inner regions of galaxies is often dominanted by the stellar component
which one can estimate using standard mass-to-light models. This data can be added
to the potential as described later in \secref{stellar mass}. By using the stellar
mass one can place a lower bound on the mass and help constrain the inner most
mass profile.
%
\begin{enumerate}
  \setcounter{enumi}{5}
  \item Point or extended mass objects can be placed in the field.
\end{enumerate}
%
\PixeLens only allows for a shear term to be added to the potential (as shown
later in \eqnref{shear}) to account for mass external to model region. This
is useful to capture the gross effects of a distant neighbour. \Glass\ also
has a shear term but additional allows for additional analytic potentials to
be included. This can be used to model substructure or multiple neighbors close
to the main lens. The substructure may have only a small effect if the lens is
a single galaxy, but if the lens is group then a potential can be added for
each of the galaxies.
%
\begin{enumerate}
  \setcounter{enumi}{5}
  \item A new uniform sampling algorithm for high dimensional spaces.
\end{enumerate}
%
At the heart of \Glass\ lies a new algorithm for sampling the high dimensional
linear space that represents the modeling solution space. This algorithm was
described and tested in \cite{}. \Glass\ is also multi-threaded allowing it to
run efficiently on many-cored machines.  The software is freely available for
download. One interesting aspect is that the input files are themselves Python
programs. This allows a large amount of sophistication in setting up a lens at
runtime and also allows \Glass\ to be used as an external library from another
user written program.

@@

The pixels only cover a finite circular area with physical radius $\Rmap$ and
pixel radius $\Rpix$ centered on the lensing galaxy. To account for any global
shearing outside this region from, e.g., a neighboring galaxy, we also add to
\eqnref{discrete potential} two shearing terms
%
\begin{equation}
\label{shear}
\gamma_1(\theta_x^2 - \theta_y^2) + 2\gamma_2\theta_x\theta_y\quad.
\end{equation}
%
We can continue adding terms to account for assumed potentials or potentials
induced from the presence of other galaxy in the field. If the stellar density
has been estimated we can use this as a lower bound where the stellar potential
is a known constant of the form \eqnref{discrete potential}.
%
%\begin{equation}
  %\psi^*(\vec\theta) = \sum_n \kappa^*_n Q_n(\vec\theta)\quad.
%\end{equation}
%

@@

\begin{table}
\begin{tabular}{cllllllll}
Galaxy & $\gamma_\star$ & $M_\star$ & $\gamma_\mathrm{DM}$ & $M_\mathrm{DM}$ & $\Rmap$ & Notes\\
\hline
AA & 1 & 4 & 0.05 & $11^{2.95}$ & 50 kpc & \\
AC & 1 & 4 & 1 & $11^2$ & 50 kpc & \\
BB & 1.5 & $2^{1.5}$ & 0.16 & $11^{2.84}$ & 50 kpc & \\
BC & 1.5 & $2^{1.5}$ & 1 & $11^2$ & 10 kpc & 
\end{tabular}
\caption{Profile parameters for the four mock galaxies. The resulting profiles only roughly follow
\eqnref{Dehnen profile} because the galaxies are triaxial. Masses are in units of $1.8\e{10}\Msun$. The scale lengths for
all lenses are $(a_\star,a_\mathrm{DM})=(2,20)$ kpc and by definition
$M_\star(a_\star) = M_\mathrm{DM}(a_\star) = 1$. $\Rmap$ is the 2d projected radius used to generate the lens configurations.}
\label{mock galaxy params}
\end{table}

\begin{figure}
\plotthree{MockGalProfile-a.pdf} {MockGalProfile-b.pdf} {MockGalProfile-c.pdf}
\caption{
\textbf{(Left)} 
Three-dimensional density of the galaxies showing the stellar, dark matter,
and total densities with dotted, dashed, and solid lines, respectively. The
grey lines show the analytic forms from \eqnref{Dehnen profile}. 
\textbf{(Middle)} 
The radially averaged two-dimensional projected density.
The critical density for lensing at $z_L=0.31$ is $\kappa_\mathrm{crit}\sim 1.8\e{9}$\Msun/kpc$^2$.
\textbf{(Right)}
The enclosed projected mass.
}
\label{mock galaxies}
\end{figure}

@@

\subsection{Stellar mass}
\label{stellar mass}
